---
# Intelligence & Cognitive Abilities (ICA) — submission checklist:
#   * Research Article. Style: APA 7th (csl: apa.csl below). Manuscript in
#     MS Word, ≤ 10,000 words including abstract, main text, references,
#     tables, figures, notes; Times New Roman or Arial 12 pt, double
#     spaced, left aligned, no line numbers (set via the reference-doc
#     on the `docx` format below); no appendices.
#   * Double-blind: the submission manuscript must be anonymized — strip
#     this author block and affiliation, the repo URL in "Data and code
#     availability", and any other self-identifying reference. A complete
#     title page goes in a SEPARATE file (reviewers do not see it).
#   * Four additional files accompany the manuscript: Title Page (authors,
#     institutions, contacts); Cover Letter (article type, IRB/conflicts,
#     corresponding author, total word count, "not under consideration
#     elsewhere", two suggested reviewers, commitment to the publication
#     fee); Abstract (separate file, ≤ 300 words); Lede / "highlights"
#     (≤ 30 words). Stubs: title_page.qmd, cover_letter.qmd, abstract.qmd,
#     lede.qmd in this directory. Queries: thomas.coyle@utsa.edu.
title: "A Model for the Distribution of SAT Total Scores"
# Anonymized for double-blind review: the author block, affiliations,
# acknowledgements, and repository URL have been removed from this file.
# Identifying information is in title_page.qmd and cover_letter.qmd,
# which are submitted separately and not seen by reviewers.
# The abstract is rendered as the first body section below (rather
# than via YAML `abstract:`) so its headline numbers can be inline-R
# expressions tied to the same posterior as the rest of the paper.
keywords: [SAT, score distributions, grouped data, sinh-arcsinh distribution, hierarchical Bayesian model, tail estimation]
bibliography: references.bib
csl: apa.csl
google-scholar: true
link-citations: true
format:
  html:
    toc: false
    number-sections: false
    fig-width: 8
    fig-height: 5
    embed-resources: true
    theme: cosmo
    fig-cap-location: bottom
    tbl-cap-location: top
    include-in-header:
      text: |
        <style>
        .quarto-figure figcaption,
        .quarto-float-tbl figcaption,
        .quarto-float-fig figcaption,
        figure > figcaption,
        caption,
        .table-caption,
        .figure-caption { text-align: left !important; }
        </style>
  docx:
    number-sections: false
    fig-width: 6.5
    fig-height: 4
    reference-doc: ica-manuscript.docx   # Times New Roman 12 pt, double spacing, left alignment, no line numbers (ICA requirement)
  typst:
    number-sections: false
    fig-width: 6.5
    fig-height: 4
    fig-cap-location: bottom
    tbl-cap-location: top
    papersize: us-letter
    margin:
      x: 1in
      y: 1in
---

```{r}
#| label: setup
#| include: false
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(purrr)
  library(ggplot2)
  library(scales)
  library(knitr)
  library(posterior)
})

# Shared figure look-and-feel.  Per-figure theme() calls below should
# only add what this default doesn't already cover (e.g., rotating
# x-axis text on a single figure).
theme_set(theme_minimal(base_size = 11) +
          theme(panel.grid.minor = element_blank(),
                legend.position  = "bottom",
                strip.text       = element_text(face = "bold", size = 10),
                axis.title       = element_text(size = 10),
                axis.text        = element_text(size = 9),
                plot.caption     = element_text(size = 8, hjust = 0)))

# Load the MI posterior (50 per-imputation .rds files + meta.rds) and
# the CSV-derived data arrays (N, mu_printed, p_obs, P_OBS).  Both
# documents/sat_model.qmd and documents/_headlines.R source this
# helper so they share a single derivation of the "data the model was
# fit to" — read straight from the per-year race CSVs, no auxiliary
# .rds dependency.  ../code/run_fit_mi.R produced the draws.
source("_inputs.R")

# Reconstruct the .mi_obj convenience bundle that the diagnostics
# chunk below references — it's just the meta fields merged onto the
# draws matrix, no extra computation.
.mi_obj <- c(list(draws = .draws_dm), .mi_meta)

# Local DRAWS wrapper providing a cmdstanr-like interface over the
# reassembled MI draws matrix (.draws_dm), so the rest of the qmd can
# call DRAWS$draws(), DRAWS$summary(), DRAWS$diagnostic_summary().
DRAWS <- list(
  draws = function(variables = NULL, format = "matrix") {
    out <- if (is.null(variables)) .draws_dm
           else subset_draws(.draws_dm, variable = variables, regex = FALSE)
    if (format == "matrix") return(out)
    if (format == "draws_df") return(as_draws_df(out))
    stop("format not supported: ", format)
  },
  summary = function(variables = NULL, ...) {
    out <- if (is.null(variables)) .draws_dm
           else subset_draws(.draws_dm, variable = variables, regex = FALSE)
    summarise_draws(out, ...)
  },
  diagnostic_summary = function() {
    list(num_divergent = .n_div_mi, num_max_treedepth = 0)
  }
)

# Undot the shared names for readability in the rest of the document.
RACES <- .RACES
YEARS <- .YEARS
SD    <- list(N = .N_arr, mu_printed = .mu_arr, p_obs = .p_obs)
P_OBS <- .P_OBS
# RACES order from the driver: White, Hispanic/Latino, Black/African American,
# Asian, Two or More Races, American Indian/Alaska Native, Native Hawaiian/
# Other Pacific Islander, No Response. Keep RACE_LABEL aligned to it.
stopifnot(identical(RACES, c("White", "Hispanic/Latino", "Black/African American",
                             "Asian", "Two or More Races",
                             "American Indian/Alaska Native",
                             "Native Hawaiian/Other Pacific Islander", "No Response")))
RACE_LABEL <- c("White", "Hispanic/Latino", "Black", "Asian",
                "Two or more", "Am. Indian", "Pacific Isl.", "No response")
BAND_LABEL <- c("400--590", "600--790", "800--990",
                "1000--1190", "1200--1390", "1400--1600")
# SAT total scores come in 10-point steps, so the band cut points sit at
# the mid-gap values (the "400-590"/"600-790" boundary is at 595, not 600).
BAND_EDGE  <- c(595, 795, 995, 1195, 1395)   # interior cut points

# Renormalise the observed band proportions to an exact simplex (as the
# Stan transformed-data block does).
P_OBS <- SD$p_obs
for (r in seq_along(RACES))
  for (y in seq_along(YEARS))
    P_OBS[r, y, ] <- P_OBS[r, y, ] / sum(P_OBS[r, y, ])

# Standard (location-0, scale-1) sinh-arcsinh CDF, and its location-scale form.
shash_F  <- function(x, eps, delta) pnorm(sinh(delta * asinh(x) - eps))
shash_Fs <- function(s, mu, sigma, eps, delta) shash_F((s - mu) / sigma, eps, delta)

# Reportable SAT total scores, and the upper edge of each one's rounding
# bin; the discrete percentile (the College Board's convention) of a pmf
# on this grid: smallest grid score whose cumulative probability >= p.
GRIDX <- seq(400, 1600, by = 10)
GRIDE <- c(GRIDX[-1] - 5, Inf)
qgrid <- function(p, cdf, x) { k <- which(cdf >= p)[1]; if (is.na(k)) x[length(x)] else x[k] }

ai <- match("Asian", RACES)
bi <- match("Black/African American", RACES)
wi <- match("White", RACES)
hi <- match("Hispanic/Latino", RACES)
y25 <- match(2025, YEARS)

# Headline numbers quoted in this manuscript's prose.  Defined in
# _headlines.R, sourced here so the abstract, the body, and the
# separately rendered abstract.qmd and cover_letter.qmd all pull from
# one place and cannot drift.  When moving this .qmd to another
# repository, copy _headlines.R along with it.
source("_headlines.R")
```

# Abstract {.unnumbered}

The College Board's SAT is the most widely used standardized test for
college and university admissions in the United States, with
approximately two million test-takers in each graduating class. The College Board does not release anonymized
individual-level scores, so a question like "what share of students
scoring at least 1500 are Black?" cannot be answered directly from the
published data. The College Board's *SAT Suite of Assessments Annual
Report* does include, for seven racial/ethnic groups across high
school graduating classes from 2018 through 2025, the percentage of
test-takers in each of six wide score bands. An eighth category ("No
response") is derived by subtraction from the published cohort band
counts. We fit a Bayesian hierarchical model to the score band
percentages using per-group censored sinh-arcsinh curves (the seven
named groups under a shared log-linear scale-trend hierarchy; the
No response group with a free per-year scale). We
validate the fit by holding out five published cohort numbers per year
(the mean, standard deviation, and three quartile scores) and an
independent fine-grained percentile reference table, none of which
is used during model fitting, and then checking that the fitted
distribution approximates them. For the graduating class of 2025, we
estimate that approximately `r blk_1500_med`% of SAT-takers
scoring 1500 or higher are Black (95% credible interval
`r blk_1500_lo`%–`r blk_1500_hi`%).

# Introduction

The College Board's SAT [@lemann1999bigtest; @collegeboard_techmanual]
is the most widely used standardized test for college and university
admissions in the United States, with approximately two million
test-takers in each graduating class. Total scores correlate around
0.8 with omnibus measures of general cognitive ability on the pre-2005
SAT [@freydetterman2004] and predict post-secondary academic
performance [@sackettkuncelhezlett2009; @kuncelhezlett2010;
@kobrinpatterson2011]. Selective colleges and merit-scholarship
programs impose explicit or de facto thresholds on it.

The College Board does not release anonymized individual-level scores.
However, it does publish, for seven racial/ethnic groups, the
percentage of that group's test-takers in each of six wide total-score
bands. A question like *what share of students scoring at least 1500
in a given graduating class are Black?*, exactly the kind of
compositional question that selective-admissions and scholarship
policy depend on, cannot be answered directly from those bands. The
1500 threshold falls inside the top band (1400--1600).
Mean group differences on the SAT and on cognitive-ability
assessments more broadly have been documented for decades
[@roth2001ethnic; @hauser2002; @reardon2011;
@reardonkalogridesshores2019], but the *shape* of each group's
upper tail, which determines who clusters above any selective cutoff,
is not visible from the published summaries alone.

The curves estimated here describe SAT *test-takers*, not the broader
U.S. 18-year-old population. The test-taker cohort is the
policy-relevant population for admissions and scholarship decisions
that condition on an SAT score. Mapping these quantities back to the
broader cohort would require a separate model of test participation
which we do not attempt here.

The College Board's annual *SAT Suite of Assessments Annual Report*
[@collegeboard_annual] gives, for each graduating class from 2018
through 2025, the number of test-takers and the mean total score on
the 400--1600 scale for each of eight racial/ethnic categories (the
seven named in the report, plus a "No response" residual we treat
as an eighth), and the percentage (integer-rounded) of the group's
test-takers in each of six wide score bands for the seven named
groups only (the No response band counts are derived by subtraction
from the published cohort band totals). The published cohort
SD, quartiles, and fine-grained percentile reference table
[@collegeboard_percentiles] are held out as external checks.

We model the latent total score for each (group, year) as a
censored sinh-arcsinh ("SHASH") random variable, with the eight
per-group curves fit jointly via Bayesian hierarchical Hamiltonian
Monte Carlo and a Dirichlet likelihood on the band percentages whose
dispersion is decoupled from group size. Methods gives the full
specification.

The fit is validated against the cohort quartiles (a test of the
model's calibration at three interior points; the model is never
shown these quartiles) and an independent
fine-grained percentile reference table [@collegeboard_percentiles].
Against the percentile reference, properly read against the integer
rounding of the published table, the model is consistent with the
published values throughout the body of the distribution and through
the headline threshold $t = 1500$, with a mild residual at $t = 1550$
(see Validation).

For the graduating class of 2025, approximately `r blk_1400_pct`%
of SAT-takers scoring 1400 or higher are Black, a share that involves
no model interpolation, since 1400 is a reported band boundary, and
follows directly from the College Board's published numbers. The corresponding share at 1500 cannot be
read off the published numbers because 1500 sits inside the top
reported band. The Black share of the 1500 or higher pool is estimated at
approximately `r blk_1500_med`%, with a 95% credible interval running
from `r blk_1500_lo`% to `r blk_1500_hi`%. The 1500 figure
involves interpolation within the top reported band, and the
implied cohort share at $t = 1500$, for the pooled 2023--2025
cohort, is consistent with the integer-rounded value in the College
Board's SAT User Percentile reference table for that same pooled
window. The model estimates an Asian-to-Black ratio of about
`r ab_1500_med`:1 among test-takers scoring 1500 or higher (95%
credible interval `r ab_1500_lo`:1 to `r ab_1500_hi`:1).

# Data

The data are drawn from the *SAT Suite of Assessments Annual Report*
[@collegeboard_annual] for the graduating classes of 2018 through 2025.
Each report tabulates, by race/ethnicity, the number of test-takers, the
mean total score (on the 400–1600 scale), and the percentage of
test-takers whose total score falls in each of six bands: 400–590,
600–790, 800–990, 1000–1190, 1200–1390, and 1400–1600. The reports
identify seven racial/ethnic categories (White, Black/African American,
Hispanic/Latino, Asian, Two or More Races, American Indian/Alaska
Native, and Native Hawaiian/Other Pacific Islander) and also report
totals for the full cohort, which is larger than the sum of the seven
because some test-takers decline to report a category.^[Throughout the
tables and figures we abbreviate the longer labels: "Black" for
Black/African American, "Two or more" for Two or More Races, "Am.
Indian" for American Indian/Alaska Native, "Pacific Isl." for Native
Hawaiian/Other Pacific Islander, and "No response" for the residual
described below.]

@tbl-pdf reproduces the seven per-group rows the report prints for the
graduating class of 2025, with the numbers shown exactly as printed
(group order in the table does not follow the report's print order).
Three features of
that raw table matter for what follows. First, the band percentages
are rounded to integers. Second, one of those rounded values is
zero (Asian in 400--590), meaning the true value is in $[0, 0.5)$
rather than that no test-taker in that group scored in that band.
Third, because of integer rounding the six bands of a given group
need not sum to exactly 100. The report also publishes test-takers
and mean total score for an eighth row, "No response", which contains
test-takers who did not disclose a race or ethnicity, but the report
does not print band percentages for that row.

```{r}
#| label: tbl-pdf
#| echo: false
#| tbl-cap: "**Per-Group Inputs from the 2025 SAT Suite Annual Report.** Test-takers ($N$), mean total score, and the percentage of each group's test-takers in each of six total-score bands, as printed by the College Board. Band percentages are integers as printed and may not sum to 100 because of rounding."
raw_pdf <- read.csv("../replicated/data/race/raw/2025.csv", stringsAsFactors = FALSE)
NAMED7 <- setdiff(RACES, "No Response")
NAMED7_LABEL <- RACE_LABEL[match(NAMED7, RACES)]
idx <- match(NAMED7, raw_pdf$race)
data.frame(
  Group   = NAMED7_LABEL,
  N       = format(raw_pdf$N[idx], big.mark = ",", trim = TRUE),
  Mean    = raw_pdf$mean[idx],
  `400--590`    = raw_pdf$b_400_590   [idx],
  `600--790`    = raw_pdf$b_600_790   [idx],
  `800--990`    = raw_pdf$b_800_990   [idx],
  `1000--1190`  = raw_pdf$b_1000_1190 [idx],
  `1200--1390`  = raw_pdf$b_1200_1390 [idx],
  `1400--1600`  = raw_pdf$b_1400_1600 [idx],
  check.names = FALSE,
  row.names = NULL
) |>
  kable(align = c("l", rep("r", 8)), row.names = FALSE)
```

The model does not fit @tbl-pdf. The
printed integer percentages are an integer-rounded representation of
a vector of true band proportions that is not observed. The fit
therefore draws $M$ datasets from the rounding-induced posterior over
plausible per-(group, year) band-proportion vectors, fits the
SHASH+Dirichlet model to each, and pools the $M$ sets of posterior
draws (multiple imputation; see *A Dirichlet Likelihood on Band
Proportions*). Each imputation proceeds in two steps. First,
for each named-group row, we draw a six-cell proportion vector
uniformly from the *rounding-box-intersected simplex*: the set of
$(\pi_1, \dots, \pi_6)$ satisfying $\pi_b \in [(k_b - 0.5)/100,\,
(k_b + 0.5)/100)$ for each band $b$ (with the lower edge clamped at
$0$ when $k_b = 0$) and $\sum_b \pi_b = 1$. Operationally, we sample
five cells independently from their intervals, set the sixth to
$1 - \sum$, and accept only if the sixth lies in its own rounding
interval; the accepted draws are within their rounding intervals
by construction and sum to one without any renormalization step.
Second, we derive the No response band counts by subtraction from the
exact published cohort band counts. The College Board publishes the
cohort *Total* row's band counts as integers (not as rounded
percentages), so the cohort row carries no rounding error of its
own, and the imputation's No response count in each band is the
cohort count minus the sum of the seven named-group counts implied
by their sampled percentages.

Because the procedure samples afresh on every imputation, there is no
single cleaned table the model fits. The pooled posterior across
$M = 50$ imputations reflects band-rounding uncertainty in addition
to whatever uncertainty the Dirichlet likelihood already captures.

Two further College Board sources are held out for validation only
and are described below where they are used (see Validation): the
published cohort summaries (mean, SD, and three quartiles per year)
and the
*SAT User Percentile* reference table
[@collegeboard_percentiles], constructed from individual scores in
ten-point steps.

```{r}
#| label: fig-data-trends
#| echo: false
#| fig-cap: "**Mean Total Score and Band-Midpoint SD by Group, 2018--2025.** Left: published mean total score. Right: SD computed by placing each band's mass at its nominal midpoint $500, 700, 900, 1100, 1300, 1500$ (the exact 10-point-grid midpoints are $495, 695, 895, 1095, 1295, 1500$; only the last falls on the round-hundred grid, because the top band 1400--1600 is 200 points wide rather than 190). The band-midpoint SD is a coarse proxy for each group's within-group SD, not the published cohort SD recovered in Validation."
#| fig-width: 9
#| fig-height: 4
MID <- c(500, 700, 900, 1100, 1300, 1500)
trends_mean <- tibble(
  race  = rep(RACE_LABEL, length(YEARS)),
  year  = rep(YEARS, each = length(RACE_LABEL)),
  panel = "Mean total score",
  value = as.vector(SD$mu_printed)
)
band_mid_sd <- sapply(seq_along(YEARS), function(yi) {
  sapply(seq_along(RACES), function(r) {
    p  <- P_OBS[r, yi, ]
    mn <- sum(p * MID)
    sqrt(sum(p * (MID - mn)^2))
  })
})
trends_sd <- tibble(
  race  = rep(RACE_LABEL, length(YEARS)),
  year  = rep(YEARS, each = length(RACE_LABEL)),
  panel = "Band-midpoint SD",
  value = as.vector(band_mid_sd)
)
trends <- bind_rows(trends_mean, trends_sd) |>
  mutate(panel = factor(panel, levels = c("Mean total score", "Band-midpoint SD")),
         race  = factor(race,  levels = RACE_LABEL))
ggplot(trends, aes(year, value, color = race, group = race)) +
  geom_line(linewidth = 0.7) + geom_point(size = 1.8) +
  facet_wrap(~ panel, scales = "free_y") +
  scale_x_continuous(breaks = YEARS, labels = function(b) substr(as.character(b), 3, 4)) +
  scale_color_brewer(palette = "Dark2", name = NULL) +
  labs(x = "Graduating-class year ('18–'25)", y = "Score points")
```

Two features of @fig-data-trends matter for the modeling choices
that follow. First, the No response *mean* drifts gently between approximately
949 and 996 across 2018–2024 and then steps to approximately 1055 in
2025, a roughly 100-point one-year jump with no parallel in any named
group, whose means move smoothly across the period. The model fits this
because every (group, year) cell has its own free mean parameter
$\theta_{r,y}$ under a weakly informative prior. No smoothness across
years is imposed on any group's mean. Second, the No response
*standard deviation* rises from approximately 211 in 2018 to
approximately 287 in 2025, a 76-point increase that dwarfs the
9–27-point rises shown by the seven named groups, and the rise is not
a smooth drift but a step pattern, with most of the jump appearing in
2020 and a further jump in 2025. The smooth log-linear scale trend the
named groups follow cannot reproduce that step structure, so the model
gives No response a free per-year scale parameter, decoupled from the
named-group hierarchy, as detailed below. No claim is made about the cause of
these movements in the No response numbers. The public record does
not distinguish a shift in who declines to disclose from a stable
population changing how it self-identifies.

```{r}
#| label: fig-band-composition
#| echo: false
#| fig-cap: "**Score-Band Composition over Time, by Group.** Each panel shows one racial/ethnic group. Within a panel, score band runs along the horizontal axis and the eight side-by-side bars within each band give the percentage of the group's test-takers in that band, by graduating-class year from 2018 (darkest) through 2025 (lightest)."
#| fig-width: 10
#| fig-height: 6.5
band_long <- purrr::map_dfr(seq_along(RACES), function(r) {
  purrr::map_dfr(seq_along(YEARS), function(yi) {
    tibble(race    = RACE_LABEL[r],
           year    = factor(YEARS[yi], levels = YEARS),
           band    = BAND_LABEL,
           percent = 100 * P_OBS[r, yi, ])
  })
}) |> mutate(race = factor(race, levels = RACE_LABEL),
             band = factor(band, levels = BAND_LABEL))
ggplot(band_long, aes(band, percent, fill = year)) +
  geom_col(position = position_dodge(width = 0.9), width = 0.85) +
  facet_wrap(~ race, ncol = 4) +
  scale_fill_viridis_d(option = "D", name = "Year") +
  labs(x = "Score band", y = "Percent of group's test-takers") +
  theme(legend.position = "bottom",
        legend.key.width = unit(0.6, "lines"),
        axis.text.x = element_text(angle = 30, hjust = 1, size = 7.5),
        strip.text = element_text(size = 9, face = "bold"))
```

@fig-band-composition makes two further patterns visible. The first
is the year-to-year *stability* of most named-group band shares: the
1400--1600 share moves by at most two or three percentage points for
every named group, with similarly narrow ranges across the bands and
across the smaller groups, and for most named-group bands the eight
side-by-side bars are close to the same height, leaving the model's
year-constant skewness and tail-weight little to fit against. The second is a steady decline in the
1000–1190 band across White, Black, and Hispanic/Latino test-takers,
mirroring the slow downward drift in those groups' means in
@fig-data-trends.

# Methods

The model has three structural components: a flexible parametric shape
for each group's score distribution, a likelihood that connects the
shape to the rounded band proportions, and a hierarchical structure
that ties the per-group parameters together. The three are described
in turn, followed by the priors and the estimation procedure.

## A Censored Sinh-Arcsinh Shape {#sec-shape}

For racial/ethnic group $r$ ($r = 1, \dots, 8$) and graduating-class
year $y$, we model an individual total score $S$ as a sinh-arcsinh
("SHASH") random variable [@jonespewsey2009], a
four-parameter generalization of the normal with separate parameters
for asymmetry and tail-thickness, used within the GAMLSS framework
for individual-level distributional modeling
[@rigbystasinopoulos2005; @stasinopoulosrigby2007]. If
$Z \sim \mathcal{N}(0, 1)$, the standardized SHASH variable is

$$
X \;=\; \sinh\!\left(\frac{\operatorname{asinh}(Z) + \epsilon}{\delta}\right),
\qquad \delta > 0,
$$ {#eq-shash}

and the score is $S = \mu_{r,y} + \sigma_{r,y} X$, with location
$\mu_{r,y}$, scale $\sigma_{r,y}$, skewness $\epsilon_r$, and
tail-weight $\delta_r$; $\epsilon = 0,\, \delta = 1$ recovers the
standard normal, $\delta < 1$ gives heavier tails and $\delta > 1$
lighter. The skewness and tail-weight are group-level and
year-constant, pooled hierarchically; the location is a free
per-(group, year) parameter under a weakly informative prior, with
the published mean playing no part. The scale varies by year through
a per-group log-linear trend plus a small per-cell residual.

Because $\sinh$ and $\operatorname{asinh}$ are monotone, the SHASH CDF
is the normal CDF $\Phi$ composed with the inverse transformation:

$$
F(s \mid \mu, \sigma, \epsilon, \delta)
  \;=\; \Phi\!\left(\sinh\!\left(\delta\,\operatorname{asinh}\!\frac{s - \mu}{\sigma} - \epsilon\right)\right).
$$ {#eq-cdf}

Censoring at the reporting range is handled by folding it into the
band boundaries. SAT total scores are reported in 10-point steps, so
the latent cut between bands "400--590" and "600--790" sits at the
midpoint of the gap, 595, and the interior cuts likewise at 795,
995, 1195, 1395. With edges
$(-\infty, 595, 795, 995, 1195, 1395, +\infty)$, the bottom band
collects mass the smooth distribution would place below 400 and the
top band mass above 1600, so the model probability that a member of
group $r$ in year $y$ falls in band $b$ is

$$
\pi_{r,y,b} \;=\; F(e_b \mid \mu_{r,y}, \sigma_{r,y}, \epsilon_r, \delta_r)
                 - F(e_{b-1} \mid \mu_{r,y}, \sigma_{r,y}, \epsilon_r, \delta_r),
$$ {#eq-bandprob}

a difference of two closed-form CDF evaluations. The same 10-point
convention applies to threshold questions later (a reported score
$\ge t$ corresponds to a latent score $\ge t-5$). The model fits on
latent scores; the censoring fold means that any latent mass the
SHASH would put above 1600 is assigned to the 1400--1600 band, and
the band probabilities are computed accordingly. Threshold queries
$P(\text{score} \ge t)$ are correspondingly latent-score queries,
which is an honest reading of the data the College Board reports
(they too truncate at 1600). For the highest-mean group (Asian,
approximately 26% of mass in the top reported band) the unrealizable
latent mass above 1600 is non-trivial; we treat within-top-band
shape as unidentified by the band data alone and revisit the
practical consequence in Composition of the High-Scoring Pool.

Of the four parameters, only the scale is given time structure. The
published cohort SD rose from approximately 204 to 235 across
2018--2025 as the SAT-taking population broadened post-SAT-optional;
a model with fixed per-group scales would sit at the all-years
average and be too narrow in recent cohorts. For each of the seven
*named* groups we let

$$
\sigma_{r,y} \;=\; \sigma_r \exp\!\bigl(\lambda_r z_y + \eta_{r,y}\bigr),
$$ {#eq-scaletrend}

with $\sigma_r$ the mid-period scale, $z_y$ the standardized year,
$\lambda_r$ a per-group log-linear growth rate (pooled toward a
common $\beta_\lambda$), and $\eta_{r,y}$ a per-cell residual around
that trend (held close to zero by a tight prior, but a genuine free
parameter so the cohort statistics inherit its uncertainty rather
than treating the parametric form as exact). The fitted
$\beta_\lambda \approx 0.03$ corresponds to scales about 10% wider in
2025 than in 2018.

The eighth group, "No response", receives a free per-year log scale
instead of the log-linear trend, for the empirical reasons given in
Data: its band-midpoint SD follows a step pattern across 2018–2025
that
a smooth trend cannot track, and forcing it into the named-group
hierarchy distorts that hierarchy.

The per-group-year location is a free parameter $\mu_{r,y}$. The
published group mean $\bar S^{\text{pub}}_{r,y}$ is not used in the fit.
It and the band percentages come from the same test-takers, so treating
the mean as a separate exact fact would double-count. Holding it out
also turns "the model recovers the published group means from the band
percentages alone" into a check comparable to those on the cohort SD
and quartiles.

The "Two or More Races" group is fit with the same single-(skewness,
tail-weight) SHASH as each named single-race group.^[The group is
structurally a mixture of distinct ancestral populations, but a
richer mixture density for this one group would not be identified by
six band totals per cell, and the empirical residuals (see
@fig-ppc-resid) do not visibly require it.]

## A Dirichlet Likelihood on Band Proportions {#sec-likelihood}

For each group and year, let
$\hat p_{r,y,\bullet} = (\hat p_{r,y,1}, \dots, \hat p_{r,y,6})$ be
the six band proportions for that imputation (each in $[0, 1]$,
summing to one by construction of the imputation procedure in Data).
We model the *observed proportions directly*, with no reconstructed
counts:

$$
\hat p_{r,y,\bullet} \;\sim\; \text{Dirichlet}\!\left(\phi_r\,\pi_{r,y,\bullet}\right),
$$ {#eq-dirichlet}

where $\pi_{r,y,\bullet}$ are the SHASH band probabilities of
@eq-bandprob and $\phi_r > 0$ is a per-group concentration parameter,
pooled hierarchically across groups (see *Hierarchical Structure*
below). The Dirichlet has mean $\pi_{r,y,\bullet}$ and component
variances $\pi_{r,y,b}(1 - \pi_{r,y,b})/(\phi_r + 1)$, which do not
depend on $N_{r,y}$. A small group is therefore not treated as orders
of magnitude noisier than a large one. The precision of every
reported proportion is governed by the per-group concentration,
fit from the data.

The group-size-invariance the Dirichlet provides addresses a mismatch
in the data. The published band percentages are *coarsened* in the
grouped-data sense of @heitjan1989 and the coarsened-data sense of
@heitjanrubin1991 and @heitjan1994, with
rounding to the nearest integer percentage point ($\pm 0.5$ percentage
points by construction). Related parametric and imputation-based
treatments of similarly coarsened or heaped data appear in
@wangheitjan2008. A multinomial likelihood on reconstructed counts would tighten the
largest group's band-proportion standard errors by a factor of about
$\sqrt{N_\text{max}/N_\text{min}} \approx 16$ relative to the smallest's,
even though the actual measurement noise --- integer rounding to the
nearest percentage point --- is the same in percentage points across
all groups; empirically, that specification gives only about 13%
coverage of nominal-90% predictive intervals. The Dirichlet decouples
dispersion from $N$ and leaves the per-group concentration to absorb
the within-group residual misfit.

Rounding noise is handled outside the Dirichlet by *multiple imputation*
[@rubin1987mi], using the rejection-sampling procedure described in
Data: each named-group row is drawn uniformly from the rounding-box-intersected
simplex, and No response is derived by subtraction from the exact
published cohort band counts. The full SHASH+Dirichlet model is fit to
each of $M = 50$ such imputed datasets and the resulting posterior
draws are pooled. Printed zeros require no separate treatment because
the rounding interval $[0, 0.5)$ is sampled natively.

Within a group, a single concentration forces the dispersion of the
wide central bands and the narrow tail bands to be governed by one
number, where in fact the rounding noise is constant in percentage
points and the Dirichlet variance scales with $\pi(1-\pi)$. A
per-band $\phi_{r,b}$ orthogonal to the per-group pooling would be
more faithful. We do not
pursue it here, and the resulting over-coverage in the tail bands is
documented in the posterior predictive check below.

## Hierarchical Structure {#sec-hierarchy}

For the seven named groups, all six per-group parameters are pooled hierarchically across
groups on the same pattern (each on a Gaussian centered at a
cross-group mean), so that the smallest groups (Native Hawaiian/Other
Pacific Islander, American Indian/Alaska Native) shrink toward the
cross-group average rather than tracking their own sampling noise.
This applies small-area estimation [@fayherriot1979;
@pfeffermann2013; @raomolina2015] at the group level. The three parameters whose natural scale is unconstrained
(the log-scale, the skewness, and the log-scale growth rate) are
pooled directly:

$$
\log\sigma_r \sim \mathcal{N}(\beta_\sigma, \tau_\sigma), \qquad
\epsilon_r \sim \mathcal{N}(\beta_\epsilon, \tau_\epsilon), \qquad
\lambda_r \sim \mathcal{N}(\beta_\lambda, \tau_\lambda).
$$ {#eq-hier}

The three positive parameters (the tail-weight, the Dirichlet
concentration, and the scale-residual SD) are pooled on the log
scale:

$$
\log\delta_r \sim \mathcal{N}(\beta_\delta, \tau_\delta), \qquad
\log\phi_r   \sim \mathcal{N}(\beta_\phi,   \tau_\phi),   \qquad
\log\tau_{\eta,r} \sim \mathcal{N}(\beta_{\tau_\eta}, \tau_{\tau_\eta}).
$$ {#eq-hier-extra}

The per-group tail-weights are *a priori* the parameters most exposed
to the within-band identifiability concern, since the within-band
shape of the top band (1400–1600) they govern is exactly what six
band totals cannot directly recover (see Discussion). In the fitted
model, the per-group tail-weight posterior medians span a narrow
range of about 0.02, all sitting just above one. The
per-group-year scale residuals themselves are
$\eta_{r,y} \sim \mathcal{N}(0, \tau_{\eta,r})$, year-level
departures from the log-linear scale trend rather than additional
structure. The "No
response" group's scale is unpooled, as described above. Its skewness
$\epsilon_{\text{NR}}$, tail-weight $\delta_{\text{NR}}$, and
concentration $\phi_{\text{NR}}$ remain in the hierarchies of
@eq-hier and @eq-hier-extra.

## Priors and Estimation {#sec-estimation}

The model's hyperparameters fall into three groups. The named-group
hierarchies on the within-group scale, skewness, and scale-growth rate
take weakly informative settings, with a Gaussian prior on each mean
and a half-normal prior on each spread:

$$
\begin{aligned}
\beta_\sigma     &\sim \mathcal{N}(\log 230,\, 0.4),  &\quad
\tau_\sigma      &\sim \mathcal{N}^{+}(0,\, 0.35), \\
\beta_\epsilon   &\sim \mathcal{N}(0,\, 0.6),         &\quad
\tau_\epsilon    &\sim \mathcal{N}^{+}(0,\, 0.8),  \\
\beta_\lambda      &\sim \mathcal{N}(0,\, 0.15),        &\quad
\tau_\lambda     &\sim \mathcal{N}^{+}(0,\, 0.06).
\end{aligned}
$$ {#eq-priors-core}

The hierarchies on the tail-weight, Dirichlet concentration, and
per-group scale-residual SD are likewise weakly informative, with the
$\beta_\delta$ and $\tau_\delta$ settings centering each $\delta_r$
near $1$:

$$
\begin{aligned}
\beta_\delta       &\sim \mathcal{N}(0,\, 0.3),           &\quad
\tau_\delta        &\sim \mathcal{N}^{+}(0,\, 0.2),  \\
\beta_\phi         &\sim \mathcal{N}(\log 300,\, 1.5),    &\quad
\tau_\phi          &\sim \mathcal{N}^{+}(0,\, 0.5),  \\
\beta_{\tau_\eta}  &\sim \mathcal{N}(\log 0.03,\, 0.6),   &\quad
\tau_{\tau_\eta}   &\sim \mathcal{N}^{+}(0,\, 0.3).
\end{aligned}
$$ {#eq-priors-aux}

The per-cell SHASH location $\mu_{r,y}$ is parameterized in terms of
the cell's *expected score* $\theta_{r,y} = \mathbb{E}[S_{r,y}]$,
which is given the prior $\theta_{r,y} \sim \mathcal{N}(1050, 300)$.
This is the *centered parameterization* of the skew-normal literature
[@azzalini1985; @azzalini2014], which removes a near-singularity in
the likelihood, applied here to the sinh-arcsinh family
[@jonespewsey2009]. The induced raw SHASH location is
$\mu_{r,y} = \theta_{r,y} - \sigma_{r,y}\,m(\epsilon_r, \delta_r)$,
with $m(\epsilon, \delta) = \mathbb{E}[X]$ computed by 24-node
Gauss–Hermite quadrature once per group per draw. The per-cell prior
mean $1050$ is the rough overall SAT mean and the spread $300$ is
wide enough to be essentially flat over the range of plausible group
means. Its only role is to keep the noisiest small group's expected
score from drifting to implausible values. In the fitted model the
posterior median per-cell expected score $\theta_{r,y}$ runs from
approximately 873 to 1248 across the eight groups and eight years
(within 0.7 prior standard deviations of the prior mean even for the
smallest groups), so the prior is essentially flat over the realized
support and does not propagate meaningfully into the posterior. For the No response group's free
per-year log scale we take
$\log\sigma_{\text{NR},y} \sim \mathcal{N}(\log 230, 0.4)$,
independent across years. The group-level hierarchy is in the
non-centered parameterization throughout.

The model is fit by Hamiltonian Monte Carlo as implemented in Stan
[@carpenter2017stan; @betancourt2017hmc], via the cmdstanr
interface [@cmdstanr2024] in R [@rcore2024]; posterior summaries and
diagnostics use the *posterior* package [@burkner2025posterior], data
manipulation uses *dplyr* [@wickham2023dplyr], and figures are
produced with *ggplot2* [@wickham2016ggplot2]. The full SHASH+Dirichlet
specification is fit independently to each of $M = 50$ imputed
datasets (drawn as described in *A Dirichlet Likelihood on Band
Proportions* above), with two chains, 1,000 warmup, and 1,000 retained
iterations per fit at `adapt_delta` = 0.95, yielding 2,000
draws per imputation and 100,000 pooled draws in total. We follow the workflow conventions of
@gabryworkflow2019 and @kruschke2014.
Convergence is assessed by the rank-normalized split-$\widehat R$
[@vehtari2021rhat] and the effective sample size. Model fit is
assessed by posterior predictive checks
[@rubin1984bayesianly; @meng1994posterior; @gelman1996ppc; @gelman2013bda]
and out-of-sample agreement with held-out summaries. We report 95%
credible intervals for the substantive headline quantities (Black
shares, the Asian-to-Black ratio, cohort summaries in the validation
figures) and 90% credible intervals for the per-group parameters in
the diagnostic tables, following common convention in each setting.
All code and data are available (see the closing note).

# Results

## Fit and Identification

```{r}
#| label: diag-summary
#| include: false
# Per-imputation diagnostics aggregated by the MI driver
max_rhat   <- .mi_obj$max_rhat_across_imp
min_bulk   <- .mi_obj$min_ess_bulk_across_imp
min_tail   <- .mi_obj$min_ess_tail_across_imp
n_div      <- .mi_obj$n_div
n_imp      <- .mi_obj$M
iters_per_imp <- nrow(.mi_obj$draws) / n_imp     # ~1,200 (2 chains x 600 samples)
total_iters   <- nrow(.mi_obj$draws)
```

Per-imputation convergence diagnostics are in @tbl-diag. Raising
the HMC adaptation target to `adapt_delta` = 0.95 from the
cmdstanr default of 0.8 cut the divergence count across the 50
imputations from a few hundred to around a dozen — well under 0.1% of
pooled iterations — without materially shifting the headline. The
worst-case $\widehat R$ and ESS [@vehtari2021rhat] across all 50
imputations sit slightly outside the conventional thresholds, with
the affected parameters concentrated at the top of the hierarchy
(the cross-group hyperparameter means) and in the No response
group. The per-cell quantities that bear on the headline
($\mu_{r,y}$, $\sigma_{r,y}$ for the named groups) have ESS bulk in
the low thousands and $\widehat R$ at most 1.01. We treat the
pooled posterior as well-resolved.

```{r}
#| label: tbl-diag
#| echo: false
#| tbl-cap: "**Convergence Diagnostics across Imputations.** Aggregated across the M = 50 imputations. Maximum rank-normalized split-$\\widehat R$ and minimum effective sample sizes are taken over per-imputation values; divergent transitions are summed."
data.frame(
  Diagnostic = c("Maximum rank-normalized split-$\\widehat R$ across imputations",
                 "Minimum bulk effective sample size across imputations",
                 "Minimum tail effective sample size across imputations",
                 sprintf("Divergent transitions (of %s pooled iterations)",
                         format(total_iters, big.mark = ","))),
  Value = c(sprintf("%.3f", max_rhat),
            format(round(min_bulk), big.mark = ","),
            format(round(min_tail), big.mark = ","),
            format(n_div, big.mark = ",")),
  check.names = FALSE
) |> kable(align = c("l", "r"))
```

@tbl-diag-shape gives per-parameter diagnostics for the cross-group
hyperparameter means and for the per-group tail-weights $\delta_r$
and skewnesses $\epsilon_r$. Across all 50 imputations the
per-group $\delta_r$ and $\epsilon_r$ have $\widehat R$ at most
around 1.01 and bulk ESS comfortably in the hundreds. The
cross-group hyperparameter means mix slowest, with the worst-case
$\widehat R$ values sitting just above 1.01 and bulk ESS in the low
hundreds in the worst-mixing imputations; that slow mixing at the
top of the hierarchy does not propagate to the per-group skewnesses
and tail-weights underneath, the parameters that carry the
substantive between-group separation.

```{r}
#| label: tbl-diag-shape
#| echo: false
#| tbl-cap: "**Per-Parameter Convergence Diagnostics.** Top-of-hierarchy hyperparameter means, plus the per-group skewnesses $\\epsilon_r$ and tail-weights $\\delta_r$. The remaining hierarchy parameters ($\\beta_\\sigma$ and all hierarchy spreads $\\tau_\\bullet$) are well-resolved in every imputation and are captured by the aggregate maximum $\\widehat R$ in @tbl-diag. Reported values are the worst case (maximum $\\widehat R$, minimum bulk ESS, minimum tail ESS) across the 50 imputations."
ROW_LBL <- c(
  beta_eps_mean     = "$\\beta_\\epsilon$",
  beta_log_delta    = "$\\beta_\\delta$",
  beta_log_phi      = "$\\beta_\\phi$",
  beta_log_tau_eta  = "$\\beta_{\\tau_\\eta}$",
  lambda_mean       = "$\\beta_\\lambda$")
for (r in seq_along(RACES))
  ROW_LBL[sprintf("epsilon[%d]", r)] <- sprintf("$\\epsilon_{\\text{%s}}$", RACE_LABEL[r])
for (r in seq_along(RACES))
  ROW_LBL[sprintf("delta[%d]", r)]   <- sprintf("$\\delta_{\\text{%s}}$", RACE_LABEL[r])
ROW_ORDER <- names(ROW_LBL)
pd <- .mi_obj$per_param_diag
pd <- pd[match(ROW_ORDER, pd$variable), ]
# Drop any rows for parameters that weren't in run_fit_mi.R's keep list
# (their diagnostics weren't saved); the table then auto-tracks the
# keep list, so adding parameters to keep just makes them appear here
# on the next refit.
keep_rows <- !is.na(pd$variable)
ROW_ORDER <- ROW_ORDER[keep_rows]
pd        <- pd[keep_rows, ]
data.frame(
  Parameter      = ROW_LBL[ROW_ORDER],
  `$\\widehat R$` = sprintf("%.3f", pd$rhat),
  `Bulk ESS`     = format(round(pd$ess_bulk), big.mark = ","),
  `Tail ESS`     = format(round(pd$ess_tail), big.mark = ","),
  check.names = FALSE,
  row.names = NULL
) |> kable(align = c("l", "r", "r", "r"), row.names = FALSE)
```

```{r}
#| include: false
# Coarse summary values used in the prose below.  The model
# parameters themselves are in @tbl-params with full precision; the
# prose uses round numbers so that small fit-to-fit drift in the
# underlying medians does not move the rendered text.
.lambda_growth_pct <- 100 * (exp(diff(range(as.numeric(scale(YEARS)))) *
                                 DRAWS$summary("lambda_mean")$median) - 1)
scale_growth_pct_5 <- sprintf("%d", 5 * round(.lambda_growth_pct / 5))
.phi_med           <- median(DRAWS$summary("phi")$median)
band_scatter_pp    <- sprintf("%.0f", 100 * sqrt(0.2 * 0.8 / .phi_med))
```

The common scale-trend rate posterior median is small and positive,
implying a scale roughly `r scale_growth_pct_5`% wider in 2025 than
in 2018, with the per-group rates in a narrow band around it and the
per-cell residuals small (per-group scale-residual SDs on the order
of one percent of the cell scale). The per-group Dirichlet
concentrations imply a band-proportion scatter of about
`r band_scatter_pp` percentage points. With rounding noise handled
upstream by the imputation layer, this residual scatter is the
year-to-year shape drift the year-constant skewness and tail-weight
cannot absorb.

@tbl-params gives the per-group parameters. Scales span
approximately 188--276 across the seven named groups. The per-group
tail-weights all sit just above one, so the tails are nearly
Gaussian for every group. What
separates the groups is the skewness: small but positive for White
(about 0.06), slightly negative for Asian (concentrated near the
1600 ceiling, with a left-trailing tail), and approximately 0.4--0.5
for the lower-mean groups (a dense lower bulk with a thin tail of
high scorers). The scale-growth rates are all small and similar. The differences in *shape* across groups
are therefore modest. The large disparities in the *composition* of
the high-scoring pool seen below are not the product of any group
having an exotic distribution, but the ordinary tail-divergence
consequence of the differences in means, which the model recovers
from the band data alone (the published means play no part).

```{r}
#| label: tbl-params
#| echo: false
#| tbl-cap: "**Per-Group Posterior Parameters for the Seven Named Groups.** Each cell shows the posterior median with the 90% credible interval in brackets. $\\sigma_r$ is the mid-period scale; $\\epsilon_r$ the (year-constant) skewness; $\\delta_r$ the (year-constant) tail-weight; $\\lambda_r$ the scale-growth rate, with $\\sigma_{r,y} = \\sigma_r\\exp(\\lambda_r z_y + \\eta_{r,y})$. The No response group is in the same skewness, tail-weight, and concentration hierarchies as the named groups (with $\\delta_{\\text{NR}} \\approx 1.13$, comparable to the named-group range), but its scale is free per year rather than parameterized by a $(\\sigma_r, \\lambda_r)$ pair (see Methods) and so it is not shown here."
gp <- function(par) {
  s <- DRAWS$summary(par)
  s$ri <- as.integer(sub(paste0("^", par, "\\[(\\d+)\\]$"), "\\1", s$variable))
  s |> arrange(ri)
}
slg <- gp("sigma_log0"); eps <- gp("epsilon"); del <- gp("delta"); lam <- gp("lambda")
nri  <- match("No Response", RACES)
named <- setdiff(seq_along(RACES), nri)
# Avoid negative-zero printing on rounded epsilon ('-0.00') by normalising.
fmt2 <- function(x) sub("^-0\\.00$", "0.00", sprintf("%.2f", x + 0))
fmt3 <- function(x) sprintf("%.3f", x)
cell_int  <- function(med, lo, hi) sprintf("%d [%d, %d]", round(med), round(lo), round(hi))
cell_fmt2 <- function(med, lo, hi) sprintf("%s [%s, %s]", fmt2(med), fmt2(lo), fmt2(hi))
cell_fmt3 <- function(med, lo, hi) sprintf("%s [%s, %s]", fmt3(med), fmt3(lo), fmt3(hi))
out <- data.frame(
  Group               = RACE_LABEL[named],
  `$\\sigma_r$`       = cell_int (exp(slg$median[named]), exp(slg$q5[named]), exp(slg$q95[named])),
  `$\\epsilon_r$`     = cell_fmt2(eps$median[named],     eps$q5[named],     eps$q95[named]),
  `$\\delta_r$`       = cell_fmt3(del$median[named],     del$q5[named],     del$q95[named]),
  `$\\lambda_r$`      = cell_fmt3(lam$median[named],     lam$q5[named],     lam$q95[named]),
  check.names = FALSE, row.names = NULL)
kable(out, align = c("l", "r", "r", "r", "r"), row.names = FALSE)
```

```{r}
#| label: draw-mats
#| include: false
# Plain numeric matrices (not the draws-matrix class) used by the
# cohort-summary and SAT User Percentile validation chunks below.
plain_mat <- function(v) { m <- DRAWS$draws(v, format = "matrix")
  matrix(as.numeric(m), nrow = nrow(m), dimnames = list(NULL, colnames(m))) }
sigry_d  <- plain_mat("sigma_ry")
eps_d    <- plain_mat("epsilon")
mu_d     <- plain_mat("mu")
thmean_d <- plain_mat("theta_mean")
del_d    <- plain_mat("delta")
nit      <- nrow(sigry_d); set.seed(20260511); keep <- sample.int(nit, 300)
sr       <- function(r, y) sprintf("sigma_ry[%d,%d]", r, y)
```

## Posterior Predictive Check

If the model is calibrated, about 90% of the observed band proportions
should lie inside their 90% posterior-predictive intervals. Drawing
replicate proportions $\tilde p_{r,y,\bullet} \sim \text{Dirichlet}(\phi\,\pi_{r,y,\bullet})$
for each of the $8 \times 8 \times 6 = 384$ cells:

```{r}
#| label: ppc-setup
#| include: false
prep <- DRAWS$summary("p_rep", ~ quantile2(.x, c(0.05, 0.5, 0.95))) |>
  mutate(idx = sub("p_rep\\[", "", variable), idx = sub("\\]", "", idx)) |>
  separate(idx, into = c("ri", "yi", "bi2"), sep = ",", convert = TRUE)
BAND_HYPHEN <- c("400-590", "600-790", "800-990", "1000-1190", "1200-1390", "1400-1600")
prep <- prep |>
  mutate(obs     = P_OBS[cbind(ri, yi, bi2)],
         in_ci   = obs >= q5 & obs <= q95,
         sd_post = (q95 - q5) / (2 * 1.645),
         z       = (obs - q50) / pmax(sd_post, 1e-9),
         group   = factor(RACE_LABEL[ri], levels = RACE_LABEL),
         year    = YEARS[yi],
         band    = factor(BAND_HYPHEN[bi2], levels = BAND_HYPHEN))
```

The aggregate coverage of the 384 cells is close to the nominal 90%.
The standardized residuals (observed minus predictive median, over
predictive SD) are centered, with spread noticeably below one — the
Dirichlet predictive intervals are *wider* than the actual residuals
warrant in many cells, particularly in the upper bands (the
1400--1600 band's z-SD $\approx 0.4$). A single per-group
concentration cannot match both the wide central bands and the
narrow tail bands at once, and the model errs toward the wide side;
a per-band concentration would tighten the tail-band intervals.

The cohort *aggregate* of the fitted band proportions agrees with
the cohort aggregate of the observed ones to within a couple of
percentage points everywhere, ruling out a bias *shared* across
groups within a year that the cell-by-cell check could miss.
@fig-ppc-resid plots the standardized residual for every (group,
year, band) cell. Most scatter around zero with spread noticeably
below one, consistent with the over-coverage just noted.

Most of the named-group breaches concentrate in the 2025 cohort: a
positive (under-predicted) cluster in the lowest band (400--590) for
the lower-mean groups, where the model under-predicts the observed
mass by about a percentage point, and a negative (over-predicted)
cluster in the adjacent 600--790 band for the same broad set, where
the model over-predicts by a comparable amount. Both patterns are
consistent with a recent-cohort shift of mass leftward in those
groups that the year-constant skewness cannot accommodate.

The No response breaches are all in the lowest band (400--590) and
include the figure's two largest residuals, both in the earlier
cohorts (2018--2019), where the observed mass sits well above what a
smooth SHASH curve fitted to the rest of that group's band
proportions will accept. The remaining No response residuals sit
closer to the named-group spread but are still slightly noisier,
consistent with the empirical volatility of the raw No response
numbers, which is also why that group is given its own free per-year
scale.

```{r}
#| label: fig-ppc-resid
#| echo: false
#| fig-cap: "**Standardized Posterior-Predictive Residuals by Group and Band.** Each point is one (group, year, band) cell. Dashed lines mark $\\pm 1.645$, the Gaussian 90% reference under a calibrated model; the asymmetric Dirichlet 90% predictive intervals reported as the coverage statistic in the text are not identical to this Gaussian cut, so the count of points outside the dashed lines need not equal the cell count outside the coverage intervals. Color encodes the score band."
#| fig-width: 8.5
#| fig-height: 4.5
ggplot(prep, aes(group, z, color = band)) +
  geom_hline(yintercept = c(-1.645, 1.645), linetype = "dashed", color = "grey55") +
  geom_hline(yintercept = 0, color = "grey25") +
  geom_point(position = position_jitter(width = 0.2, height = 0,
                                        seed = 20260524),
             alpha = 0.8, size = 1.4) +
  scale_color_viridis_d(name = NULL) +
  labs(x = NULL, y = "standardized residual") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```


## Validation {#sec-validation}

Each annual report prints, alongside the band breakdown the model is
fit to, five further cohort numbers: the Total-score mean, the
standard deviation, and the 25th, 50th, and 75th percentile scores.
None of these enters the model. The five differ in the strength of
the check they provide. The cohort mean is a weighted moment of the
same band proportions the model fits, so any model that matches the
band proportions well will recover the cohort mean close to
mechanically (up to a skewness correction inside each top band); its
posterior is identified essentially by the band proportions, and a
small recovery error is a basic consistency check rather than a
predictive test. The cohort SD is somewhat freer: the per-group
log-linear scale trend and the per-cell scale residuals give the
model flexibility to fit the cohort SD to within a few score points
without that fit being mechanically pinned by the bands. The
*quartiles*, by contrast, are a genuine out-of-sample test of the
shape in the body of the distribution: the bands constrain the
cumulative distribution at the five interior cut points (595, 795,
995, 1195, 1395), and where those cut points fall relative to the
published $p_{25}$, $p_{50}$, $p_{75}$ depends on the parametric form
between them. Across 2018–2025 that is $5 \times 8 = 40$ cohort
numbers, of which the 24 quartile cells are the operative
out-of-sample evidence.

The mean and SD are compared in score points (@fig-cohort-mean-sd).
Reported quartile *scores* live on the 10-point grid, so we compare
in the dual direction: the model's posterior rank of each published
quartile, which varies smoothly and should equal the nominal 0.25,
0.50, or 0.75 (@fig-cohort-quartile-ranks).

```{r}
#| label: cohort-stats-compute
#| include: false
muf <- function(r, y) sprintf("mu[%d,%d]", r, y)
tmf <- function(r, y) sprintf("theta_mean[%d,%d]", r, y)
pub_wide <- read.csv("../replicated/data/moments/cohort.csv") |> filter(year %in% YEARS)
cohort_stats <- map_dfr(seq_along(YEARS), function(yi) {
  w  <- SD$N[, yi] / sum(SD$N[, yi])
  pq <- pub_wide[pub_wide$year == YEARS[yi], ]
  k25 <- match(pq$p25, GRIDX); k50 <- match(pq$p50, GRIDX); k75 <- match(pq$p75, GRIDX)
  map_dfr(keep, function(it) {
    cm  <- sum(w * thmean_d[it, tmf(seq_along(RACES), yi)])
    Fhi <- sapply(seq_along(RACES), function(r)
             shash_F((GRIDE - mu_d[it, muf(r, yi)]) / sigry_d[it, sr(r, yi)], eps_d[it, r], del_d[it, r]))
    pr  <- rbind(Fhi[1, , drop = FALSE], diff(Fhi))
    pg  <- as.numeric(pr %*% w); pg <- pg / sum(pg); m <- sum(GRIDX * pg); cdf <- cumsum(pg)
    tibble(year = YEARS[yi],
           stat = factor(c("mean", "SD", "p25", "p50", "p75"),
                         levels = c("mean", "SD", "p25", "p50", "p75")),
           kind = c("score", "score", "rank", "rank", "rank"),
           value = c(cm, sqrt(sum((GRIDX - m)^2 * pg)), cdf[k25], cdf[k50], cdf[k75]))
  })
})
cohort_band <- cohort_stats |> group_by(year, stat, kind) |>
  summarize(q025 = quantile(value, .025), q50 = median(value), q975 = quantile(value, .975), .groups = "drop")
pub_pts <- bind_rows(
  pub_wide |> transmute(year, stat = "mean", kind = "score", ref = mean),
  pub_wide |> transmute(year, stat = "SD",   kind = "score", ref = sd),
  pub_wide |> transmute(year, stat = "p25",  kind = "rank",  ref = 0.25),
  pub_wide |> transmute(year, stat = "p50",  kind = "rank",  ref = 0.50),
  pub_wide |> transmute(year, stat = "p75",  kind = "rank",  ref = 0.75)) |>
  mutate(stat = factor(stat, levels = c("mean", "SD", "p25", "p50", "p75")))
```

The mean is the result that matters most. Locating each curve from
the bands alone, the model's recovered cohort mean is within a few
points of the published value every year, and per group as well.
Across the 56 named-group cells the recovered mean is within a point
or two of the published one at the median, and the published value
lies inside the model's 95% interval in nearly all of them. Only the
No response group falls clearly outside the interval, by tens of
points in some years: its mean is exactly derivable from the cohort
and named-group means by subtraction, but the band-proportion fit is
looser because of the empirical volatility shown in @fig-data-trends. The SD rises
from approximately 209 in 2018 to approximately 239 in 2025,
running about 4--5 points above the published rise from approximately
204 to 235 throughout. The small per-cell residual lets it track
the year-to-year fluctuations but not entirely (2018 and 2022 sit
just outside the posterior). These are the two years where the
published cohort SD departs furthest from a smooth trend: 2018 sits
about 6 points below the trajectory the later years define (a
plausible reading is an early-redesign test-taker cohort not yet
settled to its later composition, but the public record does not
document a specific cause), and 2022 fell by one point against an
otherwise monotone rise (the class of 2022 was the first to have
been through its junior-year SAT window with the standardized-test
requirement eliminated at almost all colleges, leaving a more
selectively self-chosen cohort). The model's per-group log-linear
scale trend with a tight residual cannot reproduce departures of
either size. The quartiles sit close to the nominal line: the
median within about a percentage point in most years (with $+1.5$
pp deviations in 2018 and 2019), $p_{25}$ within about a percentage
point in most years (rising to $+2.5$ pp in 2018), and $p_{75}$
within about a percentage point but running slightly *high* in
2024--2025 (the same recent-cohort right-skew residual). Overall, the
target lies inside the model's 95% interval in roughly nine of every
ten of the 40 cells, with the bulk of the misses in the SD and
quartile checks rather than the means. The misses themselves are
small and mostly at the interval's edge. The 24
quartile-rank checks, which exercise the shape of the cohort
distribution at three interior points that the band proportions
constrain only through the parametric form, are the operative
out-of-sample validation. The cohort mean check is a near-mechanical
consistency check of the same fit; the SD check is somewhat freer
(the within-group variances depend on the SHASH shape the model
estimates), but in practice the bands constrain it tightly enough
that it tracks the published cohort SD to within a few points.

```{r}
#| label: fig-cohort-mean-sd
#| echo: false
#| fig-cap: "**Held-Out Cross-Check of the Cohort Mean and Standard Deviation.** By graduating-class year. Neither summary was used in fitting: the model was fit only to the six per-(race, year) banded percentages, and the implied cohort mean and SD are then computed as population-weighted moments of the fitted per-group score distributions. Blue line: posterior median of the model's implied cohort value; shaded band: 95% credible interval. Black points: the College Board's published cohort value. Y-axes differ between panels."
#| fig-width: 7
#| fig-height: 3.4
cb_s <- filter(cohort_band, kind == "score")
pp_s <- filter(pub_pts,     kind == "score")
ggplot() +
  geom_ribbon(data = cb_s, aes(year, ymin = q025, ymax = q975), fill = "#2c7fb8", alpha = 0.2) +
  geom_line(  data = cb_s, aes(year, q50), color = "#2c7fb8", linewidth = 0.8) +
  geom_point( data = pp_s, aes(year, ref), color = "grey10", size = 1.9) +
  facet_wrap(~ stat, nrow = 1, scales = "free_y",
             labeller = as_labeller(c(mean = "Mean", SD = "SD"))) +
  scale_x_continuous(breaks = YEARS, labels = function(b) substr(as.character(b), 3, 4)) +
  labs(x = "Graduating-class year ('18–'25)", y = "Score points")
```

```{r}
#| label: fig-cohort-quartile-ranks
#| echo: false
#| fig-cap: "**Held-Out Cross-Check of the Published Cohort Quartile Scores.** By graduating-class year. For each year the College Board reports the score at the cohort's 25th, 50th, and 75th percentile. We plug each of those three reported scores into the model's posterior cumulative distribution to get the percentile the model itself assigns to that score (the model's 'implied rank'). A perfectly calibrated model places the published 25th-percentile score at percentile 25, the median score at 50, and the 75th-percentile score at 75. The vertical axis shows the model's implied rank minus the nominal value; zero (dashed) is perfect agreement, identical across panels so the three checks line up visually. Blue line: posterior median; shaded band: 95% credible interval. None of the three published quartiles was used in fitting."
#| fig-width: 7
#| fig-height: 3.4
TARG <- c(p25 = 25, p50 = 50, p75 = 75)
LBL  <- c(p25 = "25th percentile", p50 = "50th percentile", p75 = "75th percentile")
cb_q <- filter(cohort_band, kind == "rank") |>
  mutate(target   = TARG[as.character(stat)],
         dev_q50  = 100 * q50  - target,
         dev_q025 = 100 * q025 - target,
         dev_q975 = 100 * q975 - target,
         stat_lab = factor(LBL[as.character(stat)], levels = LBL))
ggplot(cb_q) +
  geom_ribbon(aes(year, ymin = dev_q025, ymax = dev_q975), fill = "#2c7fb8", alpha = 0.2) +
  geom_line(  aes(year, dev_q50), color = "#2c7fb8", linewidth = 0.8) +
  geom_hline( yintercept = 0, linetype = "dashed", color = "grey25") +
  facet_wrap(~ stat_lab, nrow = 1) +
  scale_x_continuous(breaks = YEARS, labels = function(b) substr(as.character(b), 3, 4)) +
  labs(x = "Graduating-class year ('18–'25)",
       y = "Model's implied rank minus nominal (pp)")
```

A second, finer check is independent of the band summaries entirely.
The College Board publishes a *SAT User Percentile* table giving, in
10-point steps, the cumulative share of test-takers scoring at or below
each total score in the three most recently graduated cohorts pooled.
The table is built from individual scores and is updated annually as
new cohorts are added and old ones drop out. We hold it out of the fit
and use it as an external check.

The check has two distinct parts. Below the 1400 boundary, the
comparison is primarily data-driven: the six observed band
proportions pin the model's cohort survival curve at five
band-boundary anchors, and any smooth distribution interpolating
between those anchors will track the SAT User Percentile curve to within
rounding noise. Above 1400, the comparison becomes distributional
extrapolation: the band data give a single anchor at the lower
boundary of the top band ($s = 1395$ in the latent-cut convention)
and the model's SHASH form fills in the rest. The
within-top-band shape (that is, what fraction of the 1400--1600
band's mass sits above 1500 or 1550) is not identifiable from six
band totals. The SAT User Percentile table is, in effect, that
missing within-top-band shape information made public. The model's
SHASH-based interpolation has no access to it during fitting;
comparing the model's induced cohort percentiles against the
reference is therefore the strictest external check the public
record allows.

The College Board publishes the SAT User Percentile table for five
pooled-cohort windows: 2018–2020, 2019–2021, 2021–2023, 2022–2024,
and 2023–2025 (the 2020–2022 window is not available). For the most
recent window the model's implied cohort survival is computed by
population-weighting the per-group SHASH survival functions over
the contributing years and compared in @fig-validate against the
published SAT User Percentile curve. The SAT User Percentile is reported
as an integer at each 10-point grid score, so the true
$P(\text{score}\ge t)$ lies in a one-percentage-point band (or a
one-sided band of width 0.5 at the top scores reported as "99+");
see the @fig-validate caption for details. The model's posterior
median sits inside the SAT User Percentile rounding band at essentially every
grid score through the headline threshold $t = 1500$: for 2023–2025
the model puts 3.2% (95% CI 2.9%–3.7%) above 1500 and the
SAT User Percentile band is $(2.5\%,\,3.5\%]$. The credible
interval's upper edge (3.7%) extends about 0.2 percentage points
above the band, so it is the median's position inside the band, not
full interval containment, that constitutes the check here. The
model drifts modestly
above the SAT User Percentile band from $t = 1550$ upward — by about 0.5–1.0
percentage points — before integer rounding at the top of the table
forecloses further comparison. The corresponding plots for the four
earlier windows show the same pattern and are not reproduced here.
The downstream consequence is taken up in the Discussion.

```{r}
#| label: validate-compute
#| include: false
WINDOWS <- list(
  "2018-2020" = c(2018, 2019, 2020),
  "2019-2021" = c(2019, 2020, 2021),
  "2021-2023" = c(2021, 2022, 2023),
  "2022-2024" = c(2022, 2023, 2024),
  "2023-2025" = c(2023, 2024, 2025))
multi_validate <- purrr::map_dfr(names(WINDOWS), function(wname) {
  years <- WINDOWS[[wname]]; yj <- match(years, YEARS)
  N_w <- SD$N[, yj]; tot_N <- sum(N_w)
  UP <- read.csv(sprintf("../replicated/data/percentiles/user_group_%s.csv", wname)) |>
    mutate(emp = pmax(0, 1 - user_pct / 100))
  # Rounding band for P(score >= t).  The College Board reports the
  # User Percentile as an integer K under the "at or below" convention,
  # except for the topmost scores where it reports "99+" (encoded in the
  # CSV as 99.5).  An integer K at grid score t-10 means
  # P(<=t-10)*100 in [K-0.5, K+0.5), so P(>=t) in
  # (1 - (K+0.5)/100, 1 - (K-0.5)/100] -- a one-percentage-point band.
  # A half-integer 99.5 at t-10 means P(<=t-10)*100 in [99.5, 100], so
  # P(>=t) in [0, 0.5%] -- a half-percentage-point ONE-SIDED band.
  ug_lookup <- setNames(UP$user_pct, as.character(UP$score))
  is_half <- function(x) abs(x - round(x)) > 0.25
  UP <- UP |> dplyr::mutate(
    pct_below = ug_lookup[as.character(score - 10)],
    pct_below = ifelse(is.na(pct_below), 0, pct_below),
    ug_lo = ifelse(is_half(pct_below),
                   0,
                   pmax(0, 1 - (pct_below + 0.5) / 100)),
    ug_hi = ifelse(is_half(pct_below),
                   pmin(1, 1 - (pct_below      ) / 100),
                   pmin(1, 1 - (pct_below - 0.5) / 100)))
  out <- purrr::map_dfr(UP$score, function(t) {
    vals <- purrr::map_dbl(keep, function(it) {
      tot <- 0
      for (r in seq_along(RACES))
        for (j in seq_along(yj)) {
          yy <- yj[j]
          tot <- tot + N_w[r, j] *
            (1 - shash_F((t - mu_d[it, muf(r, yy)]) / sigry_d[it, sr(r, yy)],
                         eps_d[it, r], del_d[it, r]))
        }
      tot / tot_N
    })
    tibble(score = t,
           q50 = median(vals),
           q5  = quantile(vals, 0.05),
           q95 = quantile(vals, 0.95))
  })
  out |> dplyr::mutate(window = wname) |>
    dplyr::inner_join(UP, by = "score")
})
# Median-inside-band coverage across all (window, grid score) cells:
ppc_n_in_val <- sum(multi_validate$q50 >= multi_validate$ug_lo &
                    multi_validate$q50 <= multi_validate$ug_hi)
ppc_n_tot_val <- nrow(multi_validate)

# Exact cohort P(score >= t) at every reported band lower-edge per
# window, from the published band data.  These are five hard
# data-pinned anchor points per window at scores 600, 800, 1000, 1200,
# 1400 (the lower edge of each band after the first).  They carry no
# integer-rounding uncertainty: the per-(race, year) band counts sum
# to the published cohort band counts by construction of the
# imputation.
band_edge_scores <- c(600, 800, 1000, 1200, 1400)
exact_anchors <- purrr::map_dfr(names(WINDOWS), function(wname) {
  yj <- match(WINDOWS[[wname]], YEARS)
  tot_N <- sum(SD$N[, yj])
  band_p <- vapply(seq_len(6), function(b)
    sum(sapply(yj, function(y) sum(SD$N[, y] * P_OBS[, y, b]))) / tot_N,
    numeric(1))
  cum_le_upper_edge <- cumsum(band_p)[1:5]
  tibble(window = wname,
         score  = band_edge_scores,
         p      = 1 - cum_le_upper_edge)
})
```

```{r}
#| label: fig-validate
#| echo: false
#| fig-cap: "**External Held-Out Validation against the SAT User Percentile Reference, 2023--2025 Pooled Cohort.** Grey rectangles: the range of $P(\\text{score} \\ge t)$ consistent with the integer SAT User Percentile at $t-10$ --- a one-percentage-point band, or a one-sided $[0,\\,0.5\\%]$ band at the topmost scores reported as '99+'. Solid blue: modeled cohort $P(\\text{score} \\ge t)$, posterior median of the population-weighted survival across 2023--2025. Black points: exact cohort survival at each reported band lower-boundary (600, 800, 1000, 1200, 1400). Plots for the four earlier pooled-cohort windows look similar and are omitted."
#| fig-width: 6.5
#| fig-height: 4
ggplot(dplyr::filter(multi_validate, window == "2023-2025"), aes(x = score)) +
  geom_rect(aes(xmin = score - 5, xmax = score + 5, ymin = ug_lo, ymax = ug_hi),
            fill = "grey30", color = "grey15", linewidth = 0.15, alpha = 0.55) +
  geom_line(aes(y = q50), color = "#2c7fb8", linewidth = 0.7) +
  geom_point(data = dplyr::filter(exact_anchors, window == "2023-2025"),
             aes(x = score, y = p), color = "grey15", size = 1.8) +
  scale_x_continuous(breaks = seq(400, 1600, by = 200)) +
  scale_y_continuous(labels = label_percent(accuracy = 1)) +
  labs(x = "SAT total threshold", y = "P(score ≥ t)")
```

## Composition of the High-Scoring Pool

```{r}
#| label: snap-compute
#| include: false
# Posterior quantities for fig-snapshot AND the inline numbers in the
# Applications prose below.  Defined here (before any prose refers to
# it) so that every number in this section -- in the text, the figure,
# and the caption -- pulls from the same posterior draws.
THR <- c(1400, 1450, 1500, 1550)
QLAB <- c(share = "Black share of above-threshold pool (%)", ratio = "Asian / Black ratio")
n_above_1400 <- SD$N[, y25] * P_OBS[, y25, 6]
snap <- map_dfr(THR, function(t) {
  if (t == 1400) {
    rr <- n_above_1400[ai] / n_above_1400[bi]
    pb <- 100 * n_above_1400[bi] / sum(n_above_1400)
    tibble(threshold = t, quantity = c("ratio", "share"),
           q50 = c(rr, pb), q025 = NA_real_, q975 = NA_real_)
  } else {
    d <- DRAWS$draws(sprintf("p_above_%d", t), format = "draws_df")
    counts <- sapply(seq_along(RACES), function(r) SD$N[r, y25] * d[[sprintf("p_above_%d[%d,%d]", t, r, y25)]])
    rr <- counts[, ai] / counts[, bi]
    pb <- 100 * counts[, bi] / rowSums(counts)
    tibble(threshold = rep(t, 2), quantity = c("ratio", "share"),
           q50  = c(median(rr), median(pb)),
           q025 = c(quantile(rr, .025), quantile(pb, .025)),
           q975 = c(quantile(rr, .975), quantile(pb, .975)))
  }
})
# Inline helpers used by the prose below.  Round so the rendered text
# reads like the figure (one decimal place for percentages, integer
# for ratios).
sg <- function(t, q, col = "q50") snap[snap$threshold == t & snap$quantity == q, col, drop = TRUE]
r1400 <- round(sg(1400, "ratio"))
p1400 <- round(sg(1400, "share"), 1)
r1450 <- round(sg(1450, "ratio"))
r1450_lo <- round(sg(1450, "ratio", "q025"))
r1450_hi <- round(sg(1450, "ratio", "q975"))
r1550 <- round(sg(1550, "ratio"))
r1550_lo <- round(sg(1550, "ratio", "q025"))
r1550_hi <- round(sg(1550, "ratio", "q975"))
```

@fig-snapshot shows the Black share and the Asian-to-Black ratio of
test-takers scoring at or above four thresholds
(1400, 1450, 1500, 1550) for the 2025 graduating class cohort. At
the 1400 boundary the
values follow directly from the College Board's published numbers (subject
to their $\pm 0.5$ percentage-point integer rounding, which is
non-trivial when the printed share is itself near $1\%$): Black
students are approximately `r p1400`% of the above-1400 pool, with
about `r r1400` Asian test-takers above 1400 per Black. The
higher thresholds fall inside the top band, where the model
interpolates. The credible intervals are wide (the 1450+ ratio is
about `r r1450` with 95% CI roughly `r r1450_lo`–`r r1450_hi`, the
1550+ ratio about `r r1550` with CI `r r1550_lo`–`r r1550_hi`). We
stop at 1550 because the model's $P(\text{latent} \ge t)$ includes
mass past the 1600 ceiling that no reported score can realize: the
effect is small at moderate thresholds (the SAT User Percentile
check passes at 1500), grows toward the ceiling, and dominates as
$t$ approaches 1600. The medians trace the familiar tail-divergence
pattern; the figure's substantive content is the width of the
intervals, which would be tighter under any model that treated the
rounded inputs as exact counts.

```{r}
#| label: fig-snapshot
#| echo: false
#| fig-cap: "**Principal Tail Quantities for the 2025 Graduating Class.** Left: Black share of all test-takers above each threshold. Right: Asian-to-Black ratio of test-takers at or above each threshold. At $t = 1400$ the values follow directly from the College Board's published numbers and are shown as point estimates without a credible interval, since the model is not used at this threshold; the only uncertainty is the $\\pm 0.5$ percentage-point integer-rounding noise on each per-race share in the 1400--1600 band, which is a deterministic bound (not a credible interval) and is discussed in the text. The three higher thresholds fall inside the top band where the model interpolates: points show posterior medians and bars show 95% credible intervals."
#| fig-width: 7.5
#| fig-height: 3.8
snap$quantity <- factor(QLAB[snap$quantity], levels = QLAB)
ggplot(snap, aes(x = threshold, y = q50)) +
  geom_errorbar(aes(ymin = q025, ymax = q975), width = 14, linewidth = 0.6,
                color = "#2c7fb8", na.rm = TRUE) +
  geom_point(size = 2.1, color = "#2c7fb8") +
  facet_wrap(~quantity, scales = "free_y") +
  scale_x_continuous(breaks = THR) +
  labs(x = "SAT total threshold", y = NULL)
```

# Discussion

This paper presents a Bayesian hierarchical model that recovers a
continuous per-group score distribution from coarsely banded summary
data. We then apply it to the College Board's reports of SAT total
scores for the graduating classes of 2018 through 2025, using the
fitted distributions to answer composition and tail-quantity
questions about the 2025 cohort. Four modeling choices
underpin the result. A sinh-arcsinh shape gives separate skewness
and tail-weight parameters with a closed-form CDF that evaluates in
constant time inside the likelihood. A free per-cell location holds
the published group mean out of the fit, which both avoids
double-counting it as a separate fact about the same students and
makes it a held-out quantity that the model has to recover. A
Dirichlet likelihood on the observed proportions provides a
group-size-invariant overdispersion that the per-group concentration
parameter fits to the actual residuals, in contrast to the sampling
noise assumed by a multinomial on reconstructed counts. And a
partial-pooling hierarchy across groups stabilizes the skewness,
tail-weight, and concentration parameters for the smaller cohorts
(Native Hawaiian/Other Pacific Islander, American Indian/Alaska
Native), which would otherwise be too imprecise to support per-group
tail-quantity claims.

The contribution is the model itself, in two respects. The
sinh-arcsinh family [@jonespewsey2009] is well established in flexible
distributional modeling, principally within the GAMLSS framework
[@rigbystasinopoulos2005; @stasinopoulosrigby2007] for individual-level
data; prior work has not, to our knowledge, applied it to coarsely
banded score data with a Dirichlet likelihood whose group-size-invariant
overdispersion accommodates the rounding step and the residual misfit
of a parsimonious shape assumption. Separately, fitting a full parametric score distribution to each
(group, year) cell of the College Board's publicly released banded
summaries is, to our knowledge, new. Published work on group differences in SAT
performance has either relied on access to individual scores
[e.g. @sackettkuncelhezlett2009; @kuncelhezlett2010;
@kobrinpatterson2011] or has stayed at the level of published
moments and aggregate gap estimates [e.g. @roth2001ethnic;
@hauser2002]. The methods literature on inferring
score gaps from coarse public data [@reardon2011;
@reardonkalogridesshores2019] uses multilevel structures to recover
mean gaps with calibrated uncertainty from heaped or banded
state-accountability test data, but has not paired a parametric
within-group distribution with a group-size-invariant Dirichlet
likelihood on the proportions themselves. The contribution here is
complementary rather than competing: within-group *shape* on
nationally aggregated SAT data, rather than between-group means on
state-accountability data. The framework proposed here is therefore the
first that allows a reader who has only the public banded reports to
ask compositional and tail-quantity questions about each racial/ethnic
group of SAT-takers separately, and to receive answers with calibrated
uncertainty.

The validation has two held-out surfaces. The five-number cohort
moments — mean, SD, and three quartile scores — are reproduced to
within a few score points (for the mean and SD) or about a
percentage point in implied rank (for the three quartiles) in
most years. The fine-grained SAT User Percentile reference
is tracked closely through the 1500 threshold, with a mild residual
at 1550. The 24 quartile cells are the genuine out-of-sample test
passed by the fit; the cohort mean is a near-mechanical consequence
of fitting the band proportions, and the cohort SD a slightly looser
corollary with some residual freedom in the within-group shape. 

The model has real limitations. All credible intervals reported on
tail quantities are conditional on the SHASH parametric form being
correct. The SHASH form interpolates the
within-1400--1600 shape that the band data cannot identify, and is
checked against the held-out SAT User Percentile reference. The
model agrees with the reference at $t = 1500$ and drifts modestly
above it at $t = 1550$ (by 0.5--1.0 percentage points in recent
windows). The cohort SD is missed in 2018 and 2022, and the
posterior rank of the published $p_{75}$ runs about a percentage
point high in 2024 and 2025.

A natural extension would be joint estimation against the SAT User
Percentile reference table, which would tighten the upper-tail fit at $t = 1550$
and above.
Beyond the SAT, the framework applies to any aggregate testing
report that publishes per-group banded distributions; any other published
quantities --- group or cohort means, SDs, percentiles --- can be
held out as additional validation checks as the SAT cohort SD and
quartiles are here. 

The paper estimates per-group *latent score distributions
conditional on test participation*, not population cognitive
distributions. Mapping the same
estimates to the broader college-bound or U.S. 18-year-old
population would require a model of who choses to take the SAT. We do not attempt that mapping here.

# Data and Code Availability

The College Board reports used here are public [@collegeboard_annual;
@collegeboard_percentiles]. Copies of those PDFs, the extraction scripts, the cleaned data,
the Stan model, and the code that produced every table and figure in
this paper are available in a public repository: \[to be supplied after blind 
peer review is complete].

# References

::: {#refs}
:::
